---
title: "Getting started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`SpectralUnmix` is designed around a simple representation of a spectral cube:
one row per spaxel, one column per wavelength channel.

## Simulate a small IFU-like dataset

```{r}
library(SpectralUnmix)

demo <- simulate_ifu_cube(nx = 14, ny = 14, n_wave = 90)
dim(demo$matrix)
```

## Fit the model

The core function estimates abundance maps and component spectra using a smooth
non-negative matrix factorization.

```{r, eval = FALSE}
fit <- spectral_unmix(demo$matrix, k = 3, niter = 250, lr = 0.03)
print(fit)
plot(fit, type = "spectra", wavelength = demo$wavelength)
plot(fit, type = "loss")
```

## Recover component maps

Once the fit is available, abundance columns can be reshaped back into image
planes for inspection.

```{r, eval = FALSE}
map1 <- component_map(fit, nx = demo$nx, ny = demo$ny, component = 1)
image(map1, main = "Component 1", xlab = "x", ylab = "y")
plot_reconstruction(fit, demo$matrix, n = 4, wavelength = demo$wavelength)
```

## Typical astronomy workflow

1. Read or simulate an IFU cube.
2. Convert it with `cube_to_matrix()`.
3. Fit `spectral_unmix()`.
4. Plot component spectra and abundance maps.
5. Compare the recovered structures with astrophysical priors or standard
   decompositions such as PCA, QR/PCA, or classical NMF.

At this stage, `SpectralUnmix` is best positioned for a methods note with one
synthetic example and one real-data application.
