---
title: "Getting Started"
---

## Simulated example

```{r, eval = FALSE}
library(SpectralUnmix)

demo <- simulate_ifu_cube(nx = 16, ny = 16, n_wave = 100)
fit <- spectral_unmix(
  demo$matrix,
  k = 3,
  lambda_smooth = 0.02,
  niter = 300,
  lr = 0.03
)
```

## Standard interface

```{r, eval = FALSE}
print(fit)
summary(fit)
basis(fit)
coef(fit)
```

## Plots

```{r, eval = FALSE}
plot(fit, type = "spectra", wavelength = demo$wavelength)
plot(fit, type = "maps", nx = demo$nx, ny = demo$ny)
plot(fit, type = "loss")
plot_reconstruction(fit, demo$matrix, n = 4, wavelength = demo$wavelength)
```

## What the outputs mean

- `fit$spatial` contains one abundance weight per spaxel and per component.
- `fit$spectra` contains the recovered component spectra.
- `fit$reconstruction` is the model approximation `A %*% S`.
- `fitted(fit, nx, ny)` and `predict(fit, type = "cube", nx, ny)` reshape that
  reconstruction back into a cube.
- `residuals(fit, x = ...)` returns data minus reconstruction.

## Real library example

```{r}
real_demo <- coelho_demo_spectra()
dim(real_demo$matrix)
real_demo$metadata
```

```{r, eval = FALSE}
fit_real <- spectral_unmix(
  real_demo$matrix,
  k = 3,
  lambda_smooth = 0.001,
  niter = 400,
  lr = 0.03
)

plot(fit_real, type = "spectra", wavelength = real_demo$wavelength)
plot_reconstruction(
  fit_real,
  real_demo$matrix,
  n = 4,
  wavelength = real_demo$wavelength
)
```

## Common failure mode

If reconstructed spectra look almost identical across the cube and only the
intensity changes, the fit has effectively collapsed toward a rank-1 solution
or toward several nearly identical components. In practice this usually means
one or more of the following:

- the input cube is dominated by a single continuum shape
- `lambda_smooth` is too large and forces spectra to become too similar
- `k` is too small, or the optimization has converged to a poor local minimum
- the data need preprocessing such as continuum normalization, masking, or
  rest-frame alignment before factorization

Useful checks are:

- plot the recovered component spectra and see whether they are nearly identical
- inspect the loss curve with `plot(fit, type = "loss")`
- retry with multiple values of `k`
- reduce `lambda_smooth`
- shift spectra to the rest frame before fitting when redshift varies

## Stellar library compression

```{r}
stellar_lib <- coelho_stellar_subset()
dim(stellar_lib$matrix)
table(stellar_lib$metadata$type)
```

```{r, eval = FALSE}
fit_lib <- spectral_unmix(
  stellar_lib$matrix,
  k = 5,
  lambda_smooth = 0.001,
  niter = 500,
  lr = 0.03
)

plot(fit_lib, type = "spectra", wavelength = stellar_lib$wavelength)
plot_reconstruction(
  fit_lib,
  stellar_lib$matrix,
  n = 6,
  wavelength = stellar_lib$wavelength
)
```
