---
title: "Getting started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`SpectralUnmix` is designed around a simple representation of a spectral cube:
one row per spaxel, one column per wavelength channel.

## Simulate a small IFU-like dataset

```{r}
library(SpectralUnmix)

demo <- simulate_ifu_cube(nx = 14, ny = 14, n_wave = 90)
dim(demo$matrix)
```

## Fit the model

The core function estimates abundance maps and component spectra using a smooth
non-negative matrix factorization.

```{r, eval = FALSE}
fit <- spectral_unmix(demo$matrix, k = 3, niter = 250, lr = 0.03)
print(fit)
plot(fit, type = "spectra", wavelength = demo$wavelength)
plot(fit, type = "loss")
```

## Recover component maps

Once the fit is available, abundance columns can be reshaped back into image
planes for inspection.

```{r, eval = FALSE}
map1 <- component_map(fit, nx = demo$nx, ny = demo$ny, component = 1)
image(map1, main = "Component 1", xlab = "x", ylab = "y")
plot_reconstruction(fit, demo$matrix, n = 4, wavelength = demo$wavelength)
```

## Real spectra example

The package also includes four real stellar spectra extracted from the Coelho
synthetic library. This is a lightweight way to show that a small number of
non-negative components can represent distinct spectral shapes from a real
library.

```{r}
real_demo <- coelho_demo_spectra()
dim(real_demo$matrix)
real_demo$metadata
```

```{r, eval = FALSE}
fit_real <- spectral_unmix(
  real_demo$matrix,
  k = 3,
  lambda_smooth = 0.001,
  niter = 400,
  lr = 0.03
)

plot(fit_real, type = "spectra", wavelength = real_demo$wavelength)
plot_reconstruction(
  fit_real,
  real_demo$matrix,
  n = 4,
  wavelength = real_demo$wavelength
)
```

## Stellar library compression example

For a larger real-data example, the package includes a curated subset of 100
stellar spectra spanning five broad stellar types. This is a better setting for
showing that a small number of non-negative components can reconstruct a real
spectral library.

```{r}
stellar_lib <- coelho_stellar_subset()
dim(stellar_lib$matrix)
table(stellar_lib$metadata$type)
```

```{r, eval = FALSE}
fit_lib <- spectral_unmix(
  stellar_lib$matrix,
  k = 5,
  lambda_smooth = 0.001,
  niter = 500,
  lr = 0.03
)

plot(fit_lib, type = "spectra", wavelength = stellar_lib$wavelength)
plot_reconstruction(
  fit_lib,
  stellar_lib$matrix,
  n = 6,
  wavelength = stellar_lib$wavelength
)
```

## Typical astronomy workflow

1. Read or simulate an IFU cube.
2. Convert it with `cube_to_matrix()`.
3. Fit `spectral_unmix()`.
4. Plot component spectra and abundance maps.
5. Compare the recovered structures with astrophysical priors or standard
   decompositions such as PCA, QR/PCA, or classical NMF.

At this stage, `SpectralUnmix` is best positioned for a methods note with one
synthetic example and one real-data application.
